Efficient pedigree recording for fast population genetics simulation

### 2017-11-16 forwards_paper.tex
overview of how things work using fwdpp

\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}

\usepackage{color}
\newcommand{\plr}[1]{{\em \color{blue} #1}}
\newcommand{\jk}[1]{{\em \color{red} #1}}

@@

The tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
Following the current best-practises [citations: apache arrow, etc] data is stored
in a columnar format.
The principal advantage of this for our purposes is that it allows for very efficient
interchange of large amounts of numerical data. In principle, this enables
zero-copy semantics, where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy
[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
data from Python into the low-level C library used to manipulate
tree sequences.
\plr{@molpopgen: note on how this works with fwdpp?}

>>>>

\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}

\usepackage{color}
\newcommand{\krt}[1]{{\em \color{green} #1}}
\newcommand{\plr}[1]{{\em \color{blue} #1}}
\newcommand{\jk}[1]{{\em \color{red} #1}}

@@

The tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
Following the current best-practises [citations: apache arrow, etc] data is stored
in a columnar format.
The principal advantage of this for our purposes is that it allows for very efficient
interchange of large amounts of numerical data. In principle, this enables
zero-copy semantics, where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy
[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
data from Python into the low-level C library used to manipulate
tree sequences.

The \msprime{} tables API is independent of the details of a particular simulation engine.  In other words, one simply
needs to fill the relevant columns and send them to \msprime{} at regular intervals for simplification.  For example,
using the \fwdpp{} C++ API \cite{Thornton2014-hx}, we fill arrays of simple structures representing the node and edge data.
Doing so requires no modification of the \fwdpp{} code base.  Rather, we simply need to book-keep the parent/offspring
labels and perform simple processing of the recombination breakpoints from each mating event.  Using \texttt{pybind11}
(\url{https://github.com/pybind/pybind11/}),
the node/edge arrays on
the C++ side are made visible, without copy, to Python as a multi-column numpy array. This array is in turn processed by
the \msprime{} tables API.  In order to create a standalone user interface, our implementation uses \texttt{pybind11} to
create a Python package extending \texttt{fwdpy11} (\url{https://github.com/molpopgen/fwdpy11}), which is a Python
package based on \fwdpp{}.  (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
API from a command-line C/C++ application via an embedded Python interpreter.)
\plr{@molpopgen: note on how this works with fwdpp?}
\krt{@plr: done, but perhaps too verbose?}
